---
title: "Duration Relational Event Models with remverse"
subtitle: "<i>A tutorial</i>"
author: ""
date: ""
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{amsmath,amssymb}
  - \usepackage{bm}
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Duration Relational Event Models with remverse}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 72, collapse = TRUE, comment = "#>")
```

---

A **Duration Relational Event Model (DuREM)** extends the standard relational event model (REM) to events that have both a *start time* and an *end time*.  Instead of a single process governing when the next event fires, there are now two linked sub-models:

- a **start model**, which describes which inactive dyad initiates a new event and when;
- an **end model**, which describes which currently active dyad terminates and when.

The two sub-models share the same parameter vector and are estimated jointly.  Statistics for each sub-model can draw on (a) the accumulated event history — weighted versions of standard remstats effects — and (b) the *current active state* of the network, i.e. the set of events that have started but not yet ended.  These active-state statistics are the contribution of `duremstats()` and cannot be expressed as weighted event history.

The remverse DuREM pipeline follows the same three-step pattern as a standard REM:

```
remify(edgelist, duration = TRUE, ...)   →  remify_durem object
  duremstats(reh, start_effects = ..., end_effects = ...)  →  remstats_durem object
    remstimate(reh, stats, ...)   →  remstimate_durem object
```

This vignette walks through each step using the `history_durem` dataset: a network of 115 directed social and work interactions among 10 actors, each with an observed start time and end time.

---

# 1  Data

## 1.1  The `history_durem` dataset

`history_durem` is derived from the `history` dataset by adding an `end` column.  Each row records one duration event:

| Column    | Description |
|-----------|-------------|
| `time`    | Start time of the event (seconds from midnight) |
| `actor1`  | Initiating actor (integer ID) |
| `actor2`  | Receiving actor (integer ID) |
| `setting` | Context of the interaction: `"work"` or `"social"` |
| `weight`  | Intrinsic event weight |
| `end`     | End time of the event (`NA` for right-censored events) |

```{r load-data}
library(remverse)   # loads remify, remstats, remstimate

data(history_durem)
head(history_durem)
```

The dataset contains a number of features that test the boundary behaviour of the pipeline:

- **End coincides with start** (rows 7, 20, 30): the end time of one event equals the start time of another — both a start and an end row appear at the same time point in the dual edgelist.
- **Simultaneous endings** (rows 23, 26): two events end at the same time point.
- **Right-censored events**: events whose end would exceed the observation window (`t = 30 000`) have `end = NA`.  They contribute to the end risk set but are never observed to terminate.

```{r data-summary}
cat(sprintf(
    "%d events: %d complete, %d right-censored\n",
    nrow(history_durem),
    sum(!is.na(history_durem$end)),
    sum( is.na(history_durem$end))
))
```

---

# 2  Step 1 — Processing with `remify()`

Pass `duration = TRUE` to tell remify that the edgelist contains duration information.  The function returns an object of class `c("remify_durem", "remify")`.

```{r remify}
reh <- remify(
    edgelist  = history_durem,
    duration  = TRUE,
    model     = "tie",
    directed  = TRUE        # actor1 initiates; end model is undirected by default
)
```

## 2.1  The `remify_durem` object

```{r print-reh}
reh
```

```{r summary-reh}
summary(reh)
```

Alongside the standard `remify` slots (`$M`, `$N`, `$D`, `$edgelist`, `$meta`, …), a `remify_durem` object carries three DuREM-specific elements:

**`$durem`** — a named list of DuREM metadata:

```{r durem-slot}
reh$durem
```

| Field          | Meaning |
|----------------|---------|
| `n_complete`   | Events with an observed end time |
| `n_censored`   | Right-censored events (`end = NA`) |
| `directed_end` | Whether the end process is directed |
| `has_who_ended`| Whether a `who_ended` column is present |
| `type_exclusive`| Whether being active in one type blocks other types |

**`$edgelist`** — the normalised start edgelist, including the `end` column.  This is the row-level record of each event and is used by `duremstats()`.

**`$edgelist_dual`** — the dual-event expansion: one `"start"` row and one `"end"` row per complete event (censored events contribute only a `"start"` row), sorted by time.  This is the sequence that the active-state C++ routines iterate over.

```{r edgelist-dual}
head(reh$edgelist_dual, 12)
```

The `duration` column on start rows equals `end − time + 1` (the `+1` ensures zero-duration events receive weight 1 under psi-weighting).  Censored start rows have `duration = 1`.  End rows record the same duration for reference.

---

# 3  Step 2 — Active-state statistics with `duremstats()`

`duremstats()` computes statistics that capture properties of the *currently active* event network — the set of dyadic events that have started but not yet ended.  These cannot be expressed as weighted event history and are the distinctive contribution of DuREM.

## 3.1  Effect functions

Effects are specified via formula objects, just like in `remstats()`.  Each effect function returns a configuration list consumed internally by `duremstats()`.

**Directed-network effects** (used in the start model when `directed = TRUE`):

| Effect | Description |
|--------|-------------|
| `activeTie()` | Is there currently an active event from actor $i$ to actor $j$? |
| `activeOutdegreeSender()` | How many active events does actor $i$ (sender) currently have as initiator? |
| `activeIndegreeReceiver()` | How many active events does actor $j$ (receiver) currently have as target? |
| `activeTotaldegreeSender()` | Total active degree of actor $i$ (as sender or receiver) |
| `activeTotaldegreeReceiver()` | Total active degree of actor $j$ |
| `activeSharedPartners_otp()` | Outgoing two-path shared partners: actors $h$ with $i \to h$ and $h \to j$ both active |
| `activeSharedPartners_itp()` | Incoming two-path version |
| `activeSharedPartners_osp()` | Outgoing shared-partner version |
| `activeSharedPartners_isp()` | Incoming shared-partner version |

**Undirected-network effects** (used in the end model when `directed_end = FALSE`):

| Effect | Description |
|--------|-------------|
| `activeTie()` | Is the dyad currently active? |
| `activeDegreeActor1()` | Active degree of actor $i$ |
| `activeDegreeActor2()` | Active degree of actor $j$ |
| `activeSharedPartners()` | Number of shared active partners |

Both families share the optional arguments:

- `scaling = c("none", "std")` — raw counts (default) or per-time-point standardisation over all dyads in the fixed risk set.
- `consider_type = c("ignore", "separate", "interact")` — how event types are taken into account (see §3.3 below).

## 3.2  Computing statistics

```{r duremstats, message=FALSE}
# Start model: is this dyad already active? how busy is the sender?
start_fx <- ~ activeTie() + activeOutdegreeSender()

# End model (undirected, default): how many active events does each actor have?
end_fx   <- ~ activeOutdegreeSender()

dstats <- duremstats(
    reh            = reh,
    start_effects  = start_fx,
    end_effects    = end_fx,
    start          = 2L,           # skip the trivially-empty first row
    display_progress = FALSE
)
```

```{r print-dstats}
dstats
```

```{r summary-dstats}
summary(dstats)
```

The returned object is of class `"remstats_durem"` and contains two 3-D arrays:

- `$start_stats` — dimensions `[M × D_start × P_start]` with effect names suffixed `.start`.
- `$end_stats` — dimensions `[M × D_end × P_end]` with effect names suffixed `.end`.

$M$ is the number of unique time points in the dual edgelist, $D$ the number of dyads in the risk set, and $P$ the number of effects.

```{r dstats-dims}
dim(dstats$start_stats)   # [M × D × P_start]
dimnames(dstats$start_stats)[[3]]

dim(dstats$end_stats)
dimnames(dstats$end_stats)[[3]]
```

### Inspecting specific statistics

The active-state statistics encode the network state *just before* each time point in the dual edgelist.  For example, we can trace how `activeTie.start` evolves for the dyad 105 → 109:

```{r inspect-stat, eval=FALSE}
# Build dyad-column lookup (mirrors the internals of duremstats)
actor_dict <- reh$meta$dictionary$actors
actor_ids  <- setNames(actor_dict$actorID - 1L, actor_dict$actorName)
riskset_mat <- convert_to_risksetMatrix(
    get_riskset(as.integer(seq_len(reh$N) - 1L), 0L, directed = TRUE),
    N = reh$N, C = 1L
)
col_105_109 <- as.integer(riskset_mat[actor_ids["105"] + 1L,
                                       actor_ids["109"] + 1L]) + 1L
# activeTie.start for dyad 105→109 across all time points
dstats$start_stats[, col_105_109, "activeTie.start"]
```

## 3.3  Event types and `consider_type`

When the edgelist contains a `type` column (here, `setting` could be renamed `type`), the `consider_type` argument controls how event types are handled in the active-state computation:

| Value | Behaviour | Output shape |
|-------|-----------|--------------|
| `"ignore"` (default) | Counts active events of any type | 1 stat, $D$ columns |
| `"separate"` | One statistic per type; only type-$c$ events contribute to the type-$c$ stat | $C$ stats, each $D$ columns |
| `"interact"` | One statistic with $D \times C$ columns; column $(i,j,c)$ reflects type-$c$ activity on dyad $(i,j)$ | 1 stat, $D \times C$ columns |

`"separate"` and `"interact"` are implemented entirely in R by filtering the dual edgelist per type; no C++ changes are required.

```{r consider-type-demo, eval=FALSE}
# Using setting as event type (rename the column first)
history_typed <- history_durem
names(history_typed)[names(history_typed) == "setting"] <- "type"
reh_typed <- remify(history_typed, duration = TRUE)

# "separate": two activeTie statistics, one per setting
dstats_sep <- duremstats(
    reh_typed,
    start_effects = ~ activeTie(consider_type = "separate")
)
# Output: activeTie.work.start and activeTie.social.start
dimnames(dstats_sep$start_stats)[[3]]

# "interact": single activeTie with D*C columns
dstats_int <- duremstats(
    reh_typed,
    start_effects = ~ activeTie(consider_type = "interact")
)
dim(dstats_int$start_stats)   # [M × (D * C) × 1]
```

---

# 4  Step 3 — Estimation with `remstimate()`

`remstimate()` dispatches on the `"remstats_durem"` class and fits the joint start/end model as a **Poisson GLM** with a log inter-event-time offset.  The stacked data frame aligns start events (with `.start` statistics) and end events (with `.end` statistics) into a single long-format design matrix; start statistics are zero for end rows and vice versa.

## 4.1  Fitting the model

```{r remstimate}
fit <- remstimate(reh = reh, stats = dstats)
```

```{r print-fit}
fit
```

## 4.2  Coefficients and log-likelihood

```{r coef}
coef(fit)
logLik(fit)
```

Each coefficient quantifies the effect of the corresponding active-state statistic on the combined start/end event rate:

- **`activeTie.start`** — the log-rate multiplier when the dyad is already active (i.e. when actor $i$ already has an active event directed at actor $j$).  A negative coefficient means that dyads already in an active event are *less* likely to initiate another event in the same direction — a form of tie saturation.

- **`activeOutdegreeSender.start`** — how the number of active outgoing events of the sender relates to starting a new event.  A negative coefficient reflects engagement: a sender who is already in many active events is less likely to initiate another.

- **`activeOutdegreeSender.end`** — how the active out-degree of the actor associated with the end event relates to termination.  A positive coefficient suggests that actors involved in more active events also end them sooner (or that high-degree actors are disproportionately responsible for terminating events).

## 4.3  Summary

```{r summary-fit}
summary(fit)
```

The `summary()` method delegates to the underlying `glm` object and reports standard Poisson GLM output: estimated coefficients, standard errors, $z$-values, and $p$-values.

## 4.4  Model comparison

Because `remstimate_durem` returns a standard `logLik` object, you can compute information criteria and compare nested models in the usual way:

```{r model-comparison, message=FALSE}
# Fit a simpler model with only activeTie in both sub-models
dstats_simple <- duremstats(
    reh,
    start_effects = ~ activeTie(),
    end_effects   = ~ activeTie(),
    start         = 2L
)
fit_simple <- remstimate(reh = reh, stats = dstats_simple)

# Compare log-likelihoods
logLik(fit)
logLik(fit_simple)

# AIC
AIC(fit)
AIC(fit_simple)
```

---

# 5  Additional topics

## 5.1  Right-censored events

Right-censored events (where `end = NA` in the edgelist) are fully supported.  They contribute a start row to the dual edgelist and appear in the active-state counts until the end of the observation window, but they never produce an observed end event.  The `$durem$n_censored` slot records how many events are censored:

```{r censored}
reh$durem$n_censored
reh$durem$n_complete
```

## 5.2  Simultaneous events and boundary coincidences

The dual edgelist may contain multiple rows at the same time point — for example, two events starting simultaneously, or one event ending at exactly the same moment that another begins.  `remify_durem` handles these cases automatically:

- **Two events ending simultaneously**: both end rows appear at the same time point.  The active-state statistics for that time point reflect the state *just before* both end.
- **End coincides with start**: an end row and a start row share a time point.  The `$edgelist_dual` is sorted so that the active state at that time point is computed consistently.

```{r simultaneous}
# Time points in the dual edgelist with more than one row
tab <- table(reh$edgelist_dual$time)
head(sort(tab[tab > 1], decreasing = TRUE), 6)
```

## 5.3  Undirected DuREM (iDynam-style)

For undirected duration events — such as face-to-face conversations — set `directed = FALSE`:

```{r undirected, eval=FALSE}
reh_ud <- remify(history_durem, duration = TRUE, directed = FALSE)

dstats_ud <- duremstats(
    reh_ud,
    start_effects = ~ activeTie() + activeDegreeActor1(),
    end_effects   = ~ activeDegreeActor1()
)

fit_ud <- remstimate(reh = reh_ud, stats = dstats_ud)
coef(fit_ud)
```

This mirrors the iDynam model of Hoffman, Block, Elmer & Stadtfeld (2020), in which the engagement effect (`activeDegreeActor1`) is the key predictor for both starting and ending face-to-face conversations.

## 5.4  Combining with history-weighted statistics

Active-state statistics from `duremstats()` can be complemented by history-weighted statistics (e.g. inertia, reciprocity) that capture the accumulated event history rather than the current active state.  When `remstats()` gains a dispatch method for `remify_durem` objects — returning a `remstats_durem` object with psi-weighted start and end statistics — the two sources of statistics will be combined into a single object before passing to `remstimate()`.  This is the intended full pipeline described in the DuREM design specification.

---

#### References

Butts, C. T. (2008). A relational event framework for social action. *Sociological Methodology*, 38(1), 155–200.

Hoffman, M., Block, P., Elmer, T., & Stadtfeld, C. (2020). A model for the dynamics of face-to-face interactions in social groups. *Network Science*, 8(S1), S4–S25.
