---
title: "Frailty Relational Event Models with GLMM"
subtitle: "<i>Random intercepts and slopes for actor heterogeneity</i>"
author: "Joris Mulder"
date: ""
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    theme: spacelab
    highlight: pygments
    code_folding: show
header-includes:
  - \usepackage{amsmath,amssymb}
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Frailty Relational Event Models with GLMM}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 72, collapse = TRUE, comment = "#>")
# Frailty fits need lme4 (declared in Suggests). If it is unavailable - e.g. a
# CRAN check run with _R_CHECK_FORCE_SUGGESTS_=false - the modelling chunks
# render code-only instead of erroring the vignette build.
run_glmm <- requireNamespace("lme4", quietly = TRUE)
```

---

Standard relational event models assume that all actors and dyads share the same set of parameters. In practice, actors differ in their baseline propensity to interact — some actors are inherently more active than others, beyond what is captured by endogenous statistics. A **frailty REM** addresses this by introducing random intercepts (and optionally random slopes) for actors, borrowing ideas from survival analysis and generalized linear mixed models (GLMMs).

In `remverse`, a GLMM is requested simply by passing a `random` formula to `remstimate()`; the convenience wrapper `remfrailty()` builds the standard sender/receiver frailty structure for you. Under the hood the stacked relational event model is fit as a Poisson GLMM (via `lme4` or `glmmTMB` for interval timing, or `coxme` for ordinal timing).

```{r load}
library(remverse)
```

---

# 1 When to use a frailty model

A frailty model is appropriate when:

- **Actor heterogeneity** is expected but not fully captured by observed covariates. Some actors may simply be more sociable or more central, independent of their network position.
- **Overdispersion** is present in the event rates — the standard Poisson model underpredicts the variance of event counts.
- **Shrinkage** is desirable — actors with few observations borrow strength from the population-level estimates rather than producing noisy individual-level estimates.

The random intercept acts as a multiplicative frailty on the event rate: actors with a positive random intercept have a higher-than-average baseline rate, while those with a negative intercept have a lower-than-average rate.

---

# 2 Data

We use the `randomREH3` event history and the `info3` actor attributes provided with `remverse`. Statistics use exponential memory decay (half-life 2000). To keep the GLMM fits fast in this vignette, we compute statistics on the last stretch of the sequence via the `first` argument (a GLMM stacks the full case-control design, which is heavier than a plain MLE fit).

```{r data}
data(randomREH3)
data(info3)

reh <- remify(edgelist = randomREH3, model = "tie", directed = TRUE)

stats <- remstats(
  reh = reh,
  tie_effects = ~ inertia(scaling = "std") + reciprocity(scaling = "std") +
                   send("age", attr_actors = info3, scaling = "std") +
                   receive("age", attr_actors = info3, scaling = "std"),
  first = 700, memory = "decay", memory_value = 2000
)
```

---

# 3 The standard model (no random effects)

As a baseline, we first estimate a standard fixed-effects-only model.

```{r standard}
fit_fixed <- remstimate(reh = reh, stats = stats)
summary(fit_fixed)
```

This model assumes all actors share the same baseline rate and the same coefficients. Any unobserved actor-level heterogeneity is absorbed into the residual.

---

# 4 Random intercepts for actors

## 4.1 Sender frailty

A random intercept for `actor1` (the sender) allows each actor to have a different baseline sending rate:

$$\log \lambda_{ij}(t) = \boldsymbol{\beta}^\top \mathbf{x}_{ij}(t) + u_i, \quad u_i \sim \mathcal{N}(0, \sigma^2_u)$$

where $u_i$ is the sender-specific random intercept. Random effects are requested via the `random` argument, which uses `glmmTMB` (default) or `lme4`.

```{r sender-frailty, message=FALSE, warning=FALSE, eval=run_glmm}
fit_sender <- remstimate(
  reh    = reh,
  stats  = stats,
  random = ~ (1 | actor1)
)
summary(fit_sender)
```

The output reports fixed-effect coefficients (comparable to the standard model) and the estimated variance of the sender random intercept ($\hat{\sigma}^2_u$). A large variance indicates substantial sender-level heterogeneity not captured by the fixed effects. (With only five actors here, the variance component is estimated from few groups and should be read as illustrative.)

## 4.2 Receiver frailty

Similarly, a random intercept for `actor2` captures receiver-level heterogeneity — some actors are more attractive as receivers regardless of the network history:

```{r receiver-frailty, message=FALSE, warning=FALSE, eval=run_glmm}
fit_receiver <- remstimate(
  reh    = reh,
  stats  = stats,
  random = ~ (1 | actor2)
)
summary(fit_receiver)
```

## 4.3 Crossed sender and receiver frailties

Both can be included simultaneously as crossed random effects. This is the standard frailty-REM specification, and `remfrailty()` builds it for you: for a directed tie model it uses `(1 | actor1) + (1 | actor2)`.

```{r crossed, message=FALSE, warning=FALSE, eval=run_glmm}
fit_crossed <- remfrailty(reh = reh, stats = stats)
summary(fit_crossed)
```

The model estimates separate variance components for sender and receiver heterogeneity, and the fixed-effect coefficients are adjusted for both sources of unobserved variation. (An explicit `remstimate(..., random = ~ (1 | actor1) + (1 | actor2))` gives the same fit.)

---

# 5 Random slopes

Beyond intercepts, we can allow the effect of an endogenous statistic to vary across actors. For example, some actors may be more habitual than others — their inertia effect is stronger — modeled with a random slope for inertia on the sender:

```{r random-slope, message=FALSE, warning=FALSE, eval=run_glmm}
stats_inertia <- remstats(
  reh = reh,
  tie_effects = ~ inertia(scaling = "std") + reciprocity(scaling = "std"),
  first = 700, memory = "decay", memory_value = 2000
)

fit_slope <- remstimate(
  reh    = reh,
  stats  = stats_inertia,
  random = ~ (1 + inertia | actor1)
)
summary(fit_slope)
```

The random effects now include a variance for the intercept, a variance for the inertia slope, and their correlation. A positive correlation would mean that actors with a high baseline sending rate also have a stronger inertia effect. Random slopes substantially increase model complexity; with small datasets the model may struggle to converge, which the `summary()` output will indicate.

---

# 6 Model comparison

We can compare models using information criteria:

```{r comparison, eval=run_glmm}
cat("Fixed-effects only: AIC =", AIC(fit_fixed),    "\n")
cat("Sender frailty:     AIC =", AIC(fit_sender),   "\n")
cat("Receiver frailty:   AIC =", AIC(fit_receiver), "\n")
cat("Crossed frailties:  AIC =", AIC(fit_crossed),  "\n")
```

A lower AIC indicates a better trade-off between fit and complexity. If a frailty model improves AIC substantially, actor-level heterogeneity is an important feature of the data.

---

# 7 Diagnostics

The GLMM diagnostics compute recall using the full linear predictor, including the best linear unbiased predictors (BLUPs) of the random effects. As throughout `remverse`, `diagnostics()` takes the fit together with the `reh` and `stats` objects it was built from.

```{r diagnostics, out.width="50%", dev=c("jpeg"), dev.args = list(bg = "white"), eval=run_glmm}
diag_crossed <- diagnostics(fit_crossed, reh, stats)
diag_crossed

plot(diag_crossed)
```

For comparison, the recall can also be computed using fixed effects only (without the random-effect predictions), by setting `use_ranef = FALSE`:

```{r diagnostics-fixed, eval=run_glmm}
diag_fixed_only <- diagnostics(fit_crossed, reh, stats, use_ranef = FALSE)
diag_fixed_only
```

The difference between the two recall values quantifies how much of the model's predictive power comes from the random effects vs. the fixed effects.

---

# 8 Interpreting random effects

The individual random effects (BLUPs) can be extracted from the fitted model to identify actors with unusually high or low activity:

```{r ranef, eval=run_glmm}
re <- lme4::ranef(fit_crossed$backend_fit)
re
```

Actors with large positive sender random effects are more active senders than average; actors with large positive receiver effects are more popular as receivers.

---

# 9 Summary

| Specification | Call | Interpretation |
|---|---|---|
| Sender frailty | `remstimate(random = ~ (1 \| actor1))` | Sender-level heterogeneity in baseline rate |
| Receiver frailty | `remstimate(random = ~ (1 \| actor2))` | Receiver-level heterogeneity (attractiveness) |
| Crossed frailties | `remfrailty(reh, stats)` | Both sender and receiver heterogeneity |
| Sender random slope | `remstimate(random = ~ (1 + inertia \| actor1))` | Actor-varying inertia effect |

The GLMM approach integrates naturally into the `remverse` pipeline. The same `remify()` and `remstats()` calls are used; only the `remstimate()` step gains a `random` argument (or is replaced by `remfrailty()`). Diagnostics and plotting work identically.


---

#### References

Juozaitiene, R., & Wit, E. C. (2024). Nodal heterogeneity can induce ghost
triadic effects in relational event models. *Psychometrika*, 89(1),
151-171. https://doi.org/10.1007/s11336-024-09952-x

Mulder, J., & Hoff, P. D. (2024). A latent variable approach for modeling
relational data with multiple receivers. *Annals of Applied Statistics*. https://doi.org/10.1214/24-AOAS1885


