---
title: "Relational Event Modeling Using 'remverse'"
subtitle: "<i>A comprehensive tutorial</i>"
author: ""
date: ""
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    theme: spacelab
    highlight: pygments
    code_folding: show
header-includes:
  - \usepackage{amsmath,amssymb}
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{The Relational Event Modeling Pipeline}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 72, collapse = TRUE, comment = "#>")
```

---

The `remverse` suite provides a three-step pipeline for relational event modeling:

1. **`remify()`** — process a raw edgelist into a standardized event history
2. **`remstats()`** — compute network statistics (endogenous and exogenous)
3. **`remstimate()`** — estimate model parameters via MLE or HMC

This vignette walks through the full pipeline for a range of modeling choices: tie-oriented vs. actor-oriented models, interval vs. ordinal timing, different risk set definitions, typed events, and a variety of endogenous and exogenous statistics. Each section builds on the same dataset so the results can be compared directly.

```{r load}
library(remverse)
```

---

# 1 Data

We use the `history` dataset (115 directed events among 10 actors) and the `info` dataset (time-varying actor attributes), both included in the `remstats` package.

```{r data}
data(history)
data(info)

head(history)
head(info)
```

Events record a `time`, a `sender` (`actor1`), a `receiver` (`actor2`), a `setting` (event type: `"work"` or `"social"`), and a `weight`. The `info` table provides `extraversion` and `agreeableness` scores measured at multiple time points for each actor.

---

# 2 Tie-oriented models {.tabset .tabset-fade .tabset-pills}

## 2.1 Basic model (interval timing)

The simplest specification: a directed tie-oriented model with interval timing (exact event times), a full risk set, and a few endogenous statistics.

```{r tie-basic}
# Step 1: process the event history
reh <- remify(edgelist = history, model = "tie", directed = TRUE)
summary(reh)

# Step 2: define effects and compute statistics
effects <- ~ inertia() + reciprocity() + indegreeSender() + outdegreeReceiver()
stats <- remstats(reh = reh, tie_effects = effects)
stats

# Step 3: estimate the model
fit <- remstimate(reh = reh, stats = stats, method = "MLE")
summary(fit)
```

The baseline intercept is automatically included for interval timing (`ordinal = FALSE`). It captures the average event rate across all dyads. The endogenous effects quantify how the accumulated event history shapes future interaction patterns: `inertia` measures the tendency to repeat past interactions, `reciprocity` captures the tendency to reciprocate received events, and the degree effects capture sender and receiver activity.

### Diagnostics

```{r tie-basic-diag, out.width="50%", dev=c("jpeg"), dev.args = list(bg = "white")}
diag <- diagnostics(object = fit, reh = reh, stats = stats)
diag

plot(x = fit, reh = reh, diagnostics = diag)
```

The `diagnostics()` function computes recall statistics: for each time point, the observed event is ranked among all competing dyads based on its predicted rate. A well-fitting model assigns high ranks to observed events. The recall summary reports the mean relative rank (1.0 = perfectly predicted) and the proportion of events in the top 5%.

---

## 2.2 Ordinal timing

When only the order of events matters (or when exact timestamps are unreliable), set `ordinal = TRUE`. The model reduces to a stratified Cox proportional-hazard model — no baseline intercept is estimated.

```{r tie-ordinal}
reh_ord <- remify(edgelist = history, model = "tie", ordinal = TRUE)

stats_ord <- remstats(reh = reh_ord, tie_effects = ~ inertia() + reciprocity())
fit_ord <- remstimate(reh = reh_ord, stats = stats_ord, method = "MLE")
summary(fit_ord)
```

The coefficients are comparable to the interval model but the intercept and inter-event-time offset are absent.

---

## 2.3 Risk set variations

### Active risk set

For sparse networks, the active risk set includes only dyads observed at least once, reducing computation substantially.

```{r tie-active}
reh_active <- remify(edgelist = history, model = "tie", riskset = "active")

cat("Full risk set:", reh$D, "dyads\n")
cat("Active risk set:", reh_active$activeD, "dyads\n")

stats_active <- remstats(reh = reh_active, tie_effects = ~ inertia() + reciprocity())
fit_active <- remstimate(reh = reh_active, stats = stats_active, method = "MLE")
summary(fit_active)
```

### Manual risk set

A manual risk set allows you to restrict the set of dyads at risk to a predefined set. Observed dyads not included in the manual specification are added automatically.

```{r tie-manual}
# Define a manual risk set: a subset of dyads
my_riskset <- data.frame(
  actor1 = c(101, 102, 103, 104, 105, 106, 107),
  actor2 = c(102, 103, 104, 105, 106, 107, 108)
)

reh_manual <- remify(
  edgelist      = history,
  model         = "tie",
  riskset       = "manual",
  manual.riskset = my_riskset
)
cat("Manual risk set:", reh_manual$activeD, "dyads\n")
```

---

## 2.4 Typed events

When events carry a type label (here, `"work"` vs. `"social"`), the `consider_type` argument controls how statistics account for event types. The risk set can optionally be extended by type (`extend_riskset_by_type = TRUE`), so that each dyad-type combination is a separate unit at risk.

```{r tie-typed}
reh_typed <- remify(
  edgelist   = history,
  model      = "tie",
  event_type = "setting"
)
```

### `consider_type = "ignore"`

Event types are not distinguished — statistics aggregate over all types.

```{r typed-ignore}
stats_ign <- remstats(
  reh = reh_typed,
  tie_effects = ~ inertia(consider_type = "ignore")
)
dimnames(stats_ign)[[3]]
```

### `consider_type = "separate"`

One statistic per type: separate counts of past `"work"` events and past `"social"` events.

```{r typed-separate}
stats_sep <- remstats(
  reh = reh_typed,
  tie_effects = ~ inertia(consider_type = "separate")
)
dimnames(stats_sep)[[3]]
```

### `consider_type = "interact"`

The statistic conditions on the type of the event at the current time point, allowing the effect of event history to depend on the triggering type.

```{r typed-interact}
stats_int <- remstats(
  reh = reh_typed,
  tie_effects = ~ inertia(consider_type = "interact")
)
dimnames(stats_int)[[3]]
```

### Estimation with typed events

```{r typed-fit}
stats_typed <- remstats(
  reh = reh_typed,
  tie_effects = ~ inertia(consider_type = "separate") +
                   reciprocity(consider_type = "separate")
)

fit_typed <- remstimate(reh = reh_typed, stats = stats_typed, method = "MLE")
summary(fit_typed)
```

The type-separated coefficients tell us whether inertia and reciprocity operate differently across event types. For instance, a stronger inertia effect for `"work"` events would suggest that work interactions are more habitual than social ones.

---

## 2.5 Exogenous statistics

Exogenous actor attributes and dyadic covariates can be included in the model alongside endogenous effects. The `info` dataset provides time-varying attributes.

### Actor-level effects

```{r exo-actor}
stats_exo <- remstats(
  reh = reh,
  tie_effects = ~ inertia() +
                   reciprocity() +
                   send("extraversion", info) +
                   receive("agreeableness", info) +
                   average("extraversion", info)
)

fit_exo <- remstimate(reh = reh, stats = stats_exo, method = "MLE")
summary(fit_exo)
```

- `send("extraversion")`: the sender's extraversion score — do extraverted actors initiate more events?
- `receive("agreeableness")`: the receiver's agreeableness — are agreeable actors more often chosen as receivers?
- `average("extraversion")`: the average extraversion of the dyad — do events flow between similarly extraverted actors?

### Interactions between endogenous and exogenous effects

Effects can be interacted to test whether endogenous tendencies (e.g., inertia) depend on actor attributes:

```{r exo-interaction}
stats_ix <- remstats(
  reh = reh,
  tie_effects = ~ inertia() +
                   send("extraversion", info) +
                   inertia():send("extraversion", info)
)

fit_ix <- remstimate(reh = reh, stats = stats_ix, method = "MLE")
summary(fit_ix)
```

A significant interaction would indicate that the tendency to repeat past interactions depends on the sender's extraversion level.

---

## 2.6 Memory types

The `memory` argument controls which past events contribute to endogenous statistics at each time point.

```{r memory}
# Full memory (default): entire event history
stats_full <- remstats(reh = reh, tie_effects = ~ inertia(), memory = "full")

# Window memory: only events within the last 500 time units
stats_win <- remstats(reh = reh, tie_effects = ~ inertia(),
                      memory = "window", memory_value = 500)

# Decay memory: exponential decay with half-life 200
stats_dec <- remstats(reh = reh, tie_effects = ~ inertia(),
                      memory = "decay", memory_value = 200)

# Compare the inertia statistic at the last time point for a single dyad
cat("Full memory:  ", stats_full[114, 1, "inertia"], "\n")
cat("Window (500): ", stats_win[114, 1, "inertia"], "\n")
cat("Decay (200):  ", stats_dec[114, 1, "inertia"], "\n")
```

Full memory accumulates all past events equally; window memory discards events older than the threshold; decay memory down-weights older events exponentially.

---

## 2.7 Case-control sampling

For large networks, computing statistics over the full risk set at every time point is expensive. Case-control sampling draws a random subset of non-event dyads as the comparison set, substantially reducing computation.

```{r sampling}
stats_sampled <- remstats(
  reh = reh,
  tie_effects = ~ inertia() + reciprocity(),
  sampling = TRUE,
  samp_num = 20,
  seed = 42
)

fit_sampled <- remstimate(reh = reh, stats = stats_sampled, method = "MLE")
summary(fit_sampled)
```

The estimates are consistent as the sample size grows but will differ slightly from the full-risk-set estimates due to sampling variability.

---

# 3 Actor-oriented models {.tabset .tabset-fade .tabset-pills}

Actor-oriented models decompose the event process into two steps: a sender rate model (which actor initiates next?) and a receiver choice model (whom does the sender choose?). This requires `directed = TRUE`.

## 3.1 Sender + receiver model

```{r actor-basic}
reh_ao <- remify(edgelist = history, model = "actor", directed = TRUE)

rate_effects <- ~ outdegreeSender() + indegreeSender()
choice_effects <- ~ inertia() + reciprocity()

stats_ao <- remstats(
  reh = reh_ao,
  sender_effects = rate_effects,
  receiver_effects = choice_effects
)
stats_ao

fit_ao <- remstimate(reh = reh_ao, stats = stats_ao, method = "MLE")
summary(fit_ao)
```

The sender model estimates include a baseline rate and the effects of out-degree and in-degree on sender activity. The receiver model estimates describe how inertia and reciprocity shape the choice of receiver, conditional on the sender.

## 3.2 Actor-oriented model with exogenous effects

```{r actor-exo}
rate_effects_exo <- ~ outdegreeSender() + send("extraversion", info)
choice_effects_exo <- ~ inertia() + reciprocity() + receive("agreeableness", info)

stats_ao_exo <- remstats(
  reh = reh_ao,
  sender_effects = rate_effects_exo,
  receiver_effects = choice_effects_exo
)

fit_ao_exo <- remstimate(reh = reh_ao, stats = stats_ao_exo, method = "MLE")
summary(fit_ao_exo)
```

## 3.3 Actor-oriented diagnostics

```{r actor-diag, out.width="50%", dev=c("jpeg"), dev.args = list(bg = "white")}
diag_ao <- diagnostics(object = fit_ao, reh = reh_ao, stats = stats_ao)
plot(x = fit_ao, reh = reh_ao, diagnostics = diag_ao)
```

---

# 4 Bayesian estimation with HMC

Both tie-oriented and actor-oriented models can be estimated with Hamiltonian Monte Carlo via `method = "HMC"`. This requires specifying the number of chains, draws, burn-in, and leap-frog parameters.

```{r hmc, message=FALSE}
fit_hmc <- remstimate(
  reh    = reh,
  stats  = stats,
  method = "HMC",
  ncores = 1L,
  nsim   = 300L,
  nchains = 2L,
  burnin  = 300L,
  L       = 100L,
  epsilon = 0.001,
  thin    = 2L,
  seed    = 42
)

summary(fit_hmc)
```

```{r hmc-diag, out.width="50%", dev=c("jpeg"), dev.args = list(bg = "white")}
diag_hmc <- diagnostics(object = fit_hmc, reh = reh, stats = stats)
plot(x = fit_hmc, reh = reh, diagnostics = diag_hmc)
```

The HMC output includes posterior means, standard deviations, and trace plots for convergence assessment.

---

# 5 Putting it all together

The table below summarizes the key modeling choices and how they map to function arguments:

| Choice | Argument | Options |
|--------|----------|---------|
| Model type | `remify(model = ...)` | `"tie"`, `"actor"` |
| Timing | `remify(ordinal = ...)` | `FALSE` (interval), `TRUE` (ordinal) |
| Risk set | `remify(riskset = ...)` | `"full"`, `"active"`, `"active_saturated"`, `"manual"` |
| Event types | `remify(event_type = ...)` | column name or `NULL` |
| Type handling | `inertia(consider_type = ...)` | `"ignore"`, `"separate"`, `"interact"` |
| Memory | `remstats(memory = ...)` | `"full"`, `"window"`, `"interval"`, `"decay"` |
| Sampling | `remstats(sampling = ...)` | `TRUE`/`FALSE` |
| Estimation | `remstimate(method = ...)` | `"MLE"`, `"HMC"` |

These choices can be freely combined — for example, an ordinal actor-oriented model with an active risk set, typed events using `consider_type = "separate"`, and Bayesian estimation via HMC.
