---
title: "Regularized Relational Event Models with GLMNET"
subtitle: "<i>Variable selection and shrinkage for REMs</i>"
author: ""
date: ""
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    theme: spacelab
    highlight: pygments
    code_folding: show
header-includes:
  - \usepackage{amsmath,amssymb}
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Regularized Relational Event Models with GLMNET}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 72, collapse = TRUE, comment = "#>")
```

---

When building relational event models, researchers often face a choice between including many network statistics to capture complex dynamics and keeping the model parsimonious to avoid overfitting. **Regularized estimation** provides a principled middle ground: fit all candidate statistics simultaneously and let the data determine which effects are important through penalized regression.

The `remverse` pipeline supports regularized estimation via `remstimate(method = "GLMNET")`, which fits the stacked relational event model using `glmnet::cv.glmnet()`. This vignette demonstrates the workflow: first fitting a rich, unregularized model to show the problem, then applying GLMNET to obtain a sparser, more robust solution.

```{r load}
library(remverse)
```

---

# 1 Data

We use the `history` dataset (115 directed events among 10 actors) and the `info` dataset (time-varying actor attributes).

```{r data}
data(history)
data(info)
history$weight <- 1

reh <- remify(edgelist = history, model = "tie")
```

---

# 2 The kitchen-sink model

We start by specifying a rich model with many endogenous and exogenous effects. This is the kind of model a researcher might consider when exploring which network mechanisms drive the event process.

```{r kitchen-sink}
effects_rich <- ~ inertia() +
                   reciprocity() +
                   indegreeSender() +
                   indegreeReceiver() +
                   outdegreeSender() +
                   outdegreeReceiver() +
                   totaldegreeSender() +
                   totaldegreeReceiver() +
                   send("extraversion", info) +
                   send("agreeableness", info) +
                   receive("extraversion", info) +
                   receive("agreeableness", info) +
                   average("extraversion", info) +
                   difference("extraversion", info)

stats_rich <- remstats(reh = reh, tie_effects = effects_rich)
dimnames(stats_rich)[[3]]
```

That gives us `r dim(stats_rich)[3]` statistics (including the baseline). With only 115 events and 90 dyads in the risk set, this is a relatively high-dimensional model.

## 2.1 Unregularized MLE

```{r mle-rich}
fit_mle <- remstimate(reh = reh, stats = stats_rich, method = "MLE")
summary(fit_mle)
```

Several things to note in the unregularized output:

- Some coefficients may have large standard errors, indicating instability.
- Degree statistics (in-degree, out-degree, total degree) are correlated with each other, potentially causing multicollinearity.
- With many parameters relative to the effective sample size, the model may overfit — performing well on the training data but poorly on held-out events.

---

# 3 Regularized estimation with GLMNET

## 3.1 Elastic net (default)

`method = "GLMNET"` fits the model with an elastic net penalty, which combines L1 (lasso) and L2 (ridge) regularization. Cross-validation selects the optimal penalty strength $\lambda$.

```{r glmnet-default}
fit_glmnet <- remstimate(reh = reh, stats = stats_rich, method = "GLMNET")
summary(fit_glmnet)
coef(fit_glmnet)
```

The regularized estimates are shrunken toward zero. Statistics with small or noisy effects are pushed to exactly zero (L1 component), effectively performing variable selection. The remaining non-zero coefficients identify the most important network mechanisms.

## 3.2 Comparing coefficients

```{r compare-coefs}
coefs_mle    <- coef(fit_mle)
coefs_glmnet <- coef(fit_glmnet)

# Align names (GLMNET drops the baseline)
shared <- intersect(names(coefs_mle), names(coefs_glmnet))

comparison <- data.frame(
  statistic = shared,
  MLE       = round(coefs_mle[shared], 4),
  GLMNET    = round(coefs_glmnet[shared], 4),
  row.names = NULL
)
comparison$selected <- comparison$GLMNET != 0
comparison
```

The `selected` column shows which effects survived regularization. Effects shrunk to zero are considered non-informative given the other effects in the model.

---

# 4 Tuning the penalty

## 4.1 Alpha: lasso vs. ridge

The `alpha` argument controls the mix between L1 (lasso, `alpha = 1`) and L2 (ridge, `alpha = 0`). The default is `alpha = 0.5` (elastic net).

```{r alpha}
# Pure lasso: maximum sparsity
fit_lasso <- remstimate(
  reh = reh, stats = stats_rich,
  method = "GLMNET", alpha = 1
)

# Pure ridge: no variable selection, only shrinkage
fit_ridge <- remstimate(
  reh = reh, stats = stats_rich,
  method = "GLMNET", alpha = 0
)

cat("Lasso non-zero:", sum(coef(fit_lasso) != 0), "of", length(coef(fit_lasso)), "\n")
cat("Ridge non-zero:", sum(coef(fit_ridge) != 0), "of", length(coef(fit_ridge)), "\n")
```

The lasso tends to select a small subset of effects, while ridge retains all effects but shrinks them. Elastic net (`alpha = 0.5`) typically offers the best balance.

## 4.2 Lambda selection

By default, `cv.glmnet` selects `lambda.min` (the $\lambda$ that minimizes cross-validated deviance). Alternatively, `lambda.1se` selects a more parsimonious model within one standard error of the minimum:

```{r lambda}
fit_1se <- remstimate(
  reh = reh, stats = stats_rich,
  method = "GLMNET", lambda_select = "1se"
)

cat("lambda.min non-zero:", sum(coef(fit_glmnet) != 0), "\n")
cat("lambda.1se non-zero:", sum(coef(fit_1se) != 0), "\n")
```

---

# 5 Diagnostics

Recall diagnostics for the regularized model use the penalized point estimates to compute linear predictors. This provides an in-sample assessment of predictive accuracy.

```{r diagnostics, out.width="50%", dev=c("jpeg"), dev.args = list(bg = "white")}
diag_mle    <- diagnostics(fit_mle)
diag_glmnet <- diagnostics(fit_glmnet)

cat("MLE recall:    ", round(diag_mle$recall$summary$mean_rel_rank, 3), "\n")
cat("GLMNET recall: ", round(diag_glmnet$recall$summary$mean_rel_rank, 3), "\n")

plot(fit_glmnet, diagnostics_object = diag_glmnet)
```

In-sample recall may be slightly lower for the regularized model (it trades a small amount of in-sample fit for better generalization). On held-out data, the regularized model often outperforms.

---

# 6 Typed events with regularization

Regularization is particularly valuable for typed event models, where `consider_type = "separate"` or `"interact"` multiplies the number of parameters by the number of event types.

```{r typed-glmnet}
reh_typed <- remify(edgelist = history, model = "tie", event_type = "setting")

stats_typed <- remstats(
  reh = reh_typed,
  tie_effects = ~ inertia(consider_type = "separate") +
                   reciprocity(consider_type = "separate") +
                   indegreeSender(consider_type = "separate") +
                   outdegreeSender(consider_type = "separate") +
                   send("extraversion", info, consider_type = "separate") +
                   receive("agreeableness", info, consider_type = "separate")
)

cat("Number of statistics:", dim(stats_typed)[3], "\n")

fit_typed_glmnet <- remstimate(
  reh = reh_typed, stats = stats_typed,
  method = "GLMNET"
)

# Which type-specific effects survived?
coefs_typed <- coef(fit_typed_glmnet)
coefs_typed[coefs_typed != 0]
```

With type-separated effects, the number of parameters doubles per type. Regularization identifies which type-specific effects carry genuine signal and which are noise.

---

# 7 Comparison with GLMM

Regularization and mixed models address different problems. GLMM captures unobserved actor-level heterogeneity through random effects; GLMNET performs variable selection and shrinkage on the fixed effects. They can be complementary: use GLMNET to identify the important effects, then fit a GLMM with those effects plus random intercepts.

```{r glmm-comparison, eval=FALSE}
# Step 1: GLMNET for variable selection
selected_stats <- names(coefs_glmnet[coefs_glmnet != 0])

# Step 2: Refit with only selected effects, adding actor random intercepts
# (This requires re-specifying the effects formula with only the selected terms)
```

---

# 8 Summary

| Aspect | MLE | GLMNET |
|--------|-----|--------|
| Estimates | Unpenalized | Shrunk toward zero |
| Variable selection | No | Yes (L1 component) |
| Multicollinearity | Problematic | Handled (L2 component) |
| Overfitting risk | Higher with many effects | Lower |
| Computation | Standard GLM | Cross-validated penalized GLM |
| Uncertainty | Standard errors available | Point estimates only |

GLMNET integrates seamlessly into the `remverse` pipeline. The same `remify()` and `remstats()` calls are used; only the `remstimate()` step changes via `method = "GLMNET"`. Diagnostics work identically to the standard model.

A natural workflow is to use GLMNET as a screening step — identifying which of many candidate effects carry signal — and then re-estimate the selected model with MLE or GLMM for proper inference (standard errors, confidence intervals, random effects).
